---
title: "Append to Fish_Tracks_Real_Time db"
author: "Michael Spear"
format:
  html:
    code-fold: true
    code-summary: Show code
resource_files:
- code/fn_get_fishtracks.R
- code/helper_functions.R
---

This Quarto appends new records from FishTracks USGS website, which displays 7 days of Real-Time acoustic tag detections at 5-minute intervals, to the IRBS database that accumulates these data with a long-term vision. Data may eventually be supplemented by a ~6-month schedule of data physically downloaded from each receiver with longer-term and/or more detailed data.

The IRBS FishTracks Real-Time database is referenced here and on Posit Connect as a DSN, but lives on the PRI Datastorm Microsoft SQL Server.

## Run log

```{r}
#| warning: FALSE
#| message: FALSE
#| error: TRUE

library(tidyverse)
library(DBI)
library(odbc)
library(here)
# source function that reads data from USGS website
source(here('code', 'helper_functions.R'))

```


```{r}
#| warning: TRUE
#| message: TRUE
#| error: TRUE
options(show.error.messages = TRUE)

# simple logging helper
log_msg <- function(msg, level = "INFO") {
  cat(sprintf("[%s] [%s] %s\n", Sys.time(), level, msg))
}


# initialize variables in case of error
rows_inserted <- NA_integer_
rows_skipped  <- NA_integer_
rows_retrieved <- NA_integer_
stations_retrieved <- NA_integer_
stations_lacking_data <- NA
compiled <- NULL
tbl_station <- NULL


tryCatch({
  
  log_msg("Job started")
  
  # Connect to db, with intuitive error handling
con <- tryCatch({

  dbConnect(odbc::odbc(), "Fish_Tracks_Real_Time")

}, error = function(e) {

  log_msg("Database connection failed", "ERROR")

  log_msg(paste("ODBC error:", e$message), "ERROR")

  stop(
    paste(
      "Failed to connect to DSN 'Fish_Tracks_Real_Time'.",
      "Possible causes:",
      "- DSN not defined on server",
      "- Network / VPN issue",
      "- SQL Server unavailable",
      "- Credentials issue",
      "\nOriginal error:", e$message
    )
  )
})
  log_msg("Connected to database")
  
  # Get stations
  tbl_station <- tbl(con, Id(schema = 'dbo', table = "station")) %>% collect()
  log_msg(sprintf("Loaded %d stations", nrow(tbl_station)))
  
  
  # Fetch data
  DateRetrieved = now()
  
  
  log_msg("Fetching FishTracks data")
  
tz_lookup <- tbl_station %>% select(station_id, tz)

compiled <- map_df(tbl_station$station_id, function(id) {
  tryCatch(
    get_fishtracks(id, tz_lookup),
    error = function(e) {
      log_msg(paste("Station failed:", id, "|", e$message), "WARN")
      return(NULL)
    }
  )
}) %>%
    mutate(
      DateRetrieved = DateRetrieved,
      # Key on the record's own timestamp, NOT the logger's Record counter:
      # the counter resets to 1 whenever the logger restarts, so Record-based
      # IDs collide with earlier rows and the append silently skips them
      # (05538010, 2026-09-23 and 2026-09-29). One record per station per
      # 5-minute poll makes station_id + TimeStamp unique by construction.
      DetectionID = paste(
        station_id,
        format(TimeStamp, '%Y%m%d%H%M%S', tz = 'UTC'),
        sep = '_'
      )
    ) %>%
    relocate(DetectionID) %>%
    # A logger occasionally polls twice within one interval (e.g. after a
    # small backward clock adjustment), giving two records with the same
    # TimeStamp; the second is usually empty. Keep the one with detections,
    # then the later Record, so a batch never carries two rows with the same
    # station + TimeStamp (the NOT EXISTS below only checks against rows
    # already in dbo.event, not within the batch). distinct() keeps the
    # first row in the current order.
    arrange(station_id, TimeStamp, desc(VRTagCount), desc(Record)) %>%
    distinct(station_id, TimeStamp, .keep_all = TRUE)
  
  
  log_msg(sprintf("Retrieved %d new records", nrow(compiled)))
  
  # Clear the staging table
  dbExecute(con, "DELETE FROM dbo.staging_event")
  # Write new data to staging table
dbWriteTable(
  con,
  Id(schema="dbo", table="staging_event"),
  value = compiled,
  append = TRUE
)
  
  
  log_msg("Wrote staging table")
  
  # Append only new rows to events table. Match on station + record time,
  # not DetectionID, so rows already stored under the old Record-based IDs
  # (pre-2026-09-30) are still recognised as duplicates.
  append_sql <- "
INSERT INTO dbo.event
SELECT *
  FROM dbo.staging_event s
WHERE NOT EXISTS (SELECT 1
  FROM dbo.event t
  WHERE t.station_id = s.station_id
    AND t.TimeStamp  = s.TimeStamp)

"
  
  rows_retrieved <- nrow(compiled)
  rows_inserted <- as.integer(dbExecute(con, append_sql))
  rows_skipped <- rows_retrieved - rows_inserted
  stations_retrieved <- length(unique(compiled$station_id))
  stations_lacking_data <- setdiff(tbl_station$station_id, unique(compiled$station_id))
  
  timestamp <- Sys.time()
  
  log_msg(sprintf("Rows retrieved: %d", rows_retrieved))
  log_msg(sprintf("Rows inserted: %d", rows_inserted))
  log_msg(sprintf("Rows skipped: %d", rows_skipped))
  log_msg(sprintf("Stations retrieved: %d/%d", stations_retrieved, length(tbl_station$station_id)))  
  log_msg(paste("Station IDs not retrieved: ", stations_lacking_data))


  
  # Log run to SQL log table
  
dbExecute(
  con,
  "INSERT INTO dbo.insert_log (run_time, rows_inserted, rows_skipped) VALUES (?, ?, ?)",
  params = list(timestamp, rows_inserted, rows_skipped)
)

  
  
  log_msg("Logged run to database")
  log_msg("Job completed successfully", "SUCCESS")
},

# Handle errors with useful logging
error = function(e){

  log_msg(paste("ERROR:", e$message), "ERROR")

  # Safe defaults so logging doesn't fail
  rows_inserted <<- if (!exists("rows_inserted")) NA_integer_ else rows_inserted
  rows_skipped  <<- if (!exists("rows_skipped"))  NA_integer_ else rows_skipped

  # Try logging error to SQL
try({
  if (exists("con") && dbIsValid(con)) {
    dbDisconnect(con)
  }

  # ✅ Reconnect fresh
  con2 <- dbConnect(odbc::odbc(), "Fish_Tracks_Real_Time")

  dbExecute(
    con2,
    "INSERT INTO dbo.insert_log (run_time, rows_inserted, rows_skipped) VALUES (?, ?, ?)",
    params = list(Sys.time(), rows_inserted, rows_skipped)
  )

  dbDisconnect(con2)

}, silent = TRUE)
  
  # IMPORTANT: print full condition so Quarto shows it
  message("Full error:")
  print(e)

  # Re-throw so Connect marks failure
  stop(e)
}
)
```

## Status

`r if (is.na(rows_inserted))"❌ Job failed" else if (rows_inserted > 0) "✅ Data inserted" else "⚠️ No new data"`

## Run Summary

-   **Run time:** `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`
-   **Stations queried:** `r nrow(tbl_station)`
-   **Stations Retrieved:** `r stations_retrieved`
-   **Station IDs lacking data:** `r stations_lacking_data`
-   **Rows retrieved:** `r nrow(compiled)`
-   **Rows inserted:** `r rows_inserted`
-   **Rows skipped (duplicates):** `r nrow(compiled) - rows_inserted`

## Rows inserted by run

```{r}
log_data <- tryCatch({
  con_report <- dbConnect(odbc::odbc(), "Fish_Tracks_Real_Time")
  on.exit(dbDisconnect(con_report), add = TRUE)
  tbl(con_report, Id(schema = "dbo", table = "insert_log")) %>% collect()
}, error = function(e) {
  log_msg(paste("Failed to read insert_log:", e$message), "ERROR")
  NULL
})

if (!is.null(log_data)) {
  ggplot(log_data, aes(run_time, rows_inserted)) +
    geom_line() +
    geom_point() +
    labs(title = "Rows Inserted per Run", x = "Run Time", y = "Rows Inserted")
}

# Disconnect
dbDisconnect(con)

```
