---
title: "Append to Fish_Tracks_Real_Time db"
author: "Michael Spear"
format:
  html:
    code-fold: true
    code-summary: Show code
resource_files:
- code/fn_get_fishtracks.R
---

This Quarto appends new records from FishTracks USGS website, which displays 7 days of Real-Time acoustic tag detections at 5-minute intervals, to the IRBS database that accumulates these data with a long-term vision. Data may eventually be supplemented by a ~6-month schedule of data physically downloaded from each receiver with longer-term and/or more detailed data.

The IRBS FishTracks Real-Time database is referenced here and on Posit Connect as a DSN, but lives on the PRI Datastorm Microsoft SQL Server.

## Run log

```{r}
#| warning: FALSE
#| message: FALSE
#| error: TRUE

library(tidyverse)
library(DBI)
library(odbc)
library(here)
# source function that reads data from USGS website
source(here('code', 'fn_get_fishtracks.R'))

```


```{r}
#| warning: TRUE
#| message: TRUE
#| error: TRUE
options(show.error.messages = TRUE)

# simple logging helper
log_msg <- function(msg, level = "INFO") {
  cat(sprintf("[%s] [%s] %s\n", Sys.time(), level, msg))
}


# initialize variables in case of error
rows_inserted <- NA_integer_
rows_skipped  <- NA_integer_
rows_retrieved <- NA_integer_
stations_retrieved <- NA_integer_
stations_lacking_data <- NA
compiled <- NULL
tbl_station <- NULL


tryCatch({
  
  log_msg("Job started")
  
  # Connect to db, with intuitive error handling
con <- tryCatch({

  dbConnect(odbc::odbc(), "Fish_Tracks_Real_Time")

}, error = function(e) {

  log_msg("Database connection failed", "ERROR")

  log_msg(paste("ODBC error:", e$message), "ERROR")

  stop(
    paste(
      "Failed to connect to DSN 'Fish_Tracks_Real_Time'.",
      "Possible causes:",
      "- DSN not defined on server",
      "- Network / VPN issue",
      "- SQL Server unavailable",
      "- Credentials issue",
      "\nOriginal error:", e$message
    )
  )
})
  log_msg("Connected to database")
  
  # Get stations
  tbl_station <- tbl(con, Id(schema = 'dbo', table = "station")) %>% collect()
  log_msg(sprintf("Loaded %d stations", nrow(tbl_station)))
  
  
  # Fetch data
  DateRetrieved = now()
  
  
  log_msg("Fetching FishTracks data")
  
  compiled <- map_df(tbl_station$station_id, function(id) {
  tryCatch(
    get_fishtracks(id),
    error = function(e) {
      log_msg(paste("Station failed:", id, "|", e$message), "WARN")
      return(NULL)
    }
  )
}) %>%
    mutate(
      DateRetrieved = DateRetrieved,
      DetectionID = paste(station_id, Record, sep = '_')
    ) %>%
    relocate(DetectionID)
  
  
  log_msg(sprintf("Retrieved %d new records", nrow(compiled)))
  
  # Clear the staging table
  dbExecute(con, "DELETE FROM dbo.staging_event")
  # Write new data to staging table
dbWriteTable(
  con,
  Id(schema="dbo", table="staging_event"),
  value = compiled,
  append = TRUE
)
  
  
  log_msg("Wrote staging table")
  
  # Append only new rows to events table
  append_sql <- "
INSERT INTO dbo.event
SELECT *
  FROM dbo.staging_event s
WHERE NOT EXISTS (SELECT 1
  FROM dbo.event t
  WHERE t.DetectionID = s.DetectionID)

"
  
  rows_retrieved <- nrow(compiled)
  rows_inserted <- as.integer(dbExecute(con, append_sql))
  rows_skipped <- rows_retrieved - rows_inserted
  stations_retrieved <- length(unique(compiled$station_id))
  stations_lacking_data <- setdiff(tbl_station$station_id, unique(compiled$station_id))
  
  timestamp <- Sys.time()
  
  log_msg(sprintf("Rows retrieved: %d", rows_retrieved))
  log_msg(sprintf("Rows inserted: %d", rows_inserted))
  log_msg(sprintf("Rows skipped: %d", rows_skipped))
  log_msg(sprintf("Stations retrieved: %d/%d", stations_retrieved, length(tbl_station$station_id)))  
  log_msg(paste("Station IDs not retrieved: ", stations_lacking_data))


  
  # Log run to SQL log table
  
dbExecute(
  con,
  "INSERT INTO dbo.insert_log (run_time, rows_inserted, rows_skipped) VALUES (?, ?, ?)",
  params = list(timestamp, rows_inserted, rows_skipped)
)

  
  
  log_msg("Logged run to database")
  log_msg("Job completed successfully", "SUCCESS")
},

# Handle errors with useful logging
error = function(e){

  log_msg(paste("ERROR:", e$message), "ERROR")

  # Safe defaults so logging doesn't fail
  rows_inserted <<- if (!exists("rows_inserted")) NA_integer_ else rows_inserted
  rows_skipped  <<- if (!exists("rows_skipped"))  NA_integer_ else rows_skipped

  # Try logging error to SQL
  try({
    if (exists("con") && dbIsValid(con)) {

      dbExecute(
        con,
        "INSERT INTO dbo.insert_log (run_time, rows_inserted, rows_skipped) VALUES (?, ?, ?)",
        params = list(Sys.time(), rows_inserted, rows_skipped)
      )

      dbDisconnect(con)
    }
  }, silent = TRUE)

  # IMPORTANT: print full condition so Quarto shows it
  message("Full error:")
  print(e)

  # Re-throw so Connect marks failure
  stop(e)
}
)
```

## Status

`r if (is.na(rows_inserted))"❌ Job failed" else if (rows_inserted > 0) "✅ Data inserted" else "⚠️ No new data"`

## Run Summary

-   **Run time:** `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`
-   **Stations queried:** `r nrow(tbl_station)`
-   **Stations Retrieved:** `r stations_retrieved`
-   **Station IDs lacking data:** `r stations_lacking_data`
-   **Rows retrieved:** `r nrow(compiled)`
-   **Rows inserted:** `r rows_inserted`
-   **Rows skipped (duplicates):** `r nrow(compiled) - rows_inserted`

## Rows inserted by run

```{r}
log_data <- tbl(con, Id(schema = "dbo", table = "insert_log")) %>%
  collect()

library(ggplot2)

ggplot(log_data, aes(run_time, rows_inserted)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Rows Inserted per Run",
    x = "Run Time",
    y = "Rows Inserted"
  )

# Disconnect
dbDisconnect(con)

```
